Asteroseismology of GD358 with complex C/O core profiles

### 2016-08-06 msnew.tex
Revised, including Mike's comments
Sections 1, 5-7

\section{Astrophysical Context}
\label{intro}
White dwarfs are the end product of evolution for around 98\% of the stars in our Galaxy. 
Buried in their interiors are the records of physical processes that take place during 
earlier stages in the life of the star. Nuclear reaction rates during the core helium 
burning phase set the core composition of white dwarfs, while the relative time spent 
burning hydrogen and helium during the asymptotic-giant-branch (AGB) phase and mass-loss 
episodes determine the thickness of the helium layer \citep{Lawlor06,Althaus05}. Helium 
atmosphere white dwarfs (DBs) comprise roughly 20\% of the population of field white 
dwarfs, with most of the remaining 80\% consisting of their hydrogen atmosphere (DA) 
cousins. The bifurcation into two spectral classes is thought to occur during post-AGB 
evolution when, in some cases, a very late thermal pulse burns off the residual hydrogen 
in the envelope, producing a nearly pure helium atmosphere \citep{Iben83}. Such objects 
then return to the white dwarf cooling track as PG 1159 stars, which are 
widely believed to be the precursors of most DB white dwarfs. DBs are found to pulsate at 
effective temperatures ranging between 21,000 K and 28,000 K \citep{Beauchamp99, Castanheira05}.

The subject of this paper, GD358 (V777 Her) is the brightest ($m_{\rm v}=13.7$) and best 
studied helium atmosphere white dwarf pulsator. It is located near the red edge of the 
instability strip, with a spectroscopic temperature of \teff\ $=24000\pm500$ ~K and \logg\ $=7.8$ 
\citep{Koester2013}. GD358's pulsation spectrum contains a series of independent 
modes, many with complex frequency structure.  Models involving magnetic fields and oblique rotation 
have been proposed to explain that structure \citep{Montgomery10}.

Since the 2006 Whole Earth Telescope (WET) run reported in \citet{Provencal09}, 
we have maintained an active observing program of GD358 in an effort to identify additional
frequencies in this complex star. These new observations have identified additional 
periods in GD358's pulsation spectrum, bringing the total known modes to 15. Thirteen of 
these modes belong to a consecutive $\ell=1$ sequence, the longest sequence observed 
in a DBV.

Paradoxally, among the DBVs with enough detected periods to be fitted 
asteroseismically, GD358 is the only one that has not been analyzed using the complex 
C/O profiles adapted and parameterized from stellar evolution calculations
(e.g. {}\citet{Salaris97,Althaus05}). The most recent fits of GD358 \citep{Metcalfe03c} were 
performed using 11 observed modes and simple models where the oxygen abundance 
drops linearly from its central value to zero. We present here a new detailed asteroseismic analysis 
of GD358, taking advantage of its long sequence of $\ell=1$ modes to better constrain 
the asteroseismic fits and define the limits of high precision white dwarf asteroseismology.

@@

To compute our models, we used the White Dwarf Evolution Code (WDEC). The WDEC evolves hot 
polytrope models from temperatures above 100{,}000~K down to the temperature of our choice. 
Models in the temperature range of interest for the present study are thermally relaxed 
solutions to the stellar structure equations. Each model we compute for our grids is the result 
of such an evolutionary sequence. The WDEC is described in detail in \citet{Lamb75} and 
\citet{Wood90}. We used smoother core composition profiles and implemented more complex profiles 
that result from stellar evolution calculations \citep{Salaris97}. We updated the envelope 
equation of state tables from those calculated by \citet{Fontaine77} to those given by 
\citet{Saumon95}. We use OPAL opacities \citep{Iglesias96} and plasmon neutrino rates 
published by \citet{Itoh96}.

@@

In our asteroseismic fits, we initially varied six parameters; the effective temperature, 
the mass and four structure parameters. There are two parameters associated with the shapes 
of the oxygen (and carbon) core composition profiles: the central oxygen abundance (\xo) and 
the edge of the homogeneous carbon and oxygen core (\qfm, as a fraction of stellar mass. 
For envelope structure, \menv\ marks the location of the base of the helium layer 
and \mhe\ the location where the helium abundance rises to $1$ (see Fig. \ref{ffit1}). 
\menv\ and \mhe\ are mass coordinates, defined as e.g. $M_{\rm env} = -\log(1 - M(r)/M_*)$, 
where $M(r)$ is the mass enclosed in radius r and $M_*$ is the stellar mass.

@@

\begin{equation}
\label{fiteq2}
\Delta P = \frac{\pi}{\ell(\ell + 1)^{1/2}}\left[\int_{r_1}^{r_2}\frac{N}{r}dr\right]^{-1},
\end{equation}

@@

\begin{figure}
\epsscale{1.0}
\plotone{ffit3.eps}
%  \includegraphics[scale=0.70,viewport=50 0 50 350]{f2.png}
%  \includegraphics[width=0.9\columnwidth]{ffit3.eps}
\caption{
Dependence of the fitness parameter on different parameters. The 5 parameters other than the one shown 
on the horizontal axis in each plot are held to fixed values corresponding to the initial best fit model (table \ref{ffit1}). The vertical dashed lines mark the ranges considered in the refined fitting. For the envelope mass, we went from \menv = -2.0 to -2.3.
 \label{ffit3}
}
\end{figure}

@@

\subsection{Validation of the Fitting Method}

@@

We first performed a fitting of the periods using the master (coarse) grid described in section \ref{grids}. This 
placed the best fit model in the appropriate region of parameter space. Then we refined our fits, using the grids 
described at the end of section \ref{periodspacing}. We are able to recover the best fit parameters adequately, with 
the all top 5 best fit models having parameters 
${\rm T_{eff}}=25000$~K, \mhe $=-2.1$, $M_{\rm env}=-5.5$, $X_o=0.50$, and $q_{\rm fm}=0.190$ and a mass ranging between  0.5730 and 0.5734 \msun. The best fit model has \sigrms $=0.31$~s. We remind the reader that 
the step sizes in the second phase of the fitting are 50~K for ${\rm T_{eff}}$, 0.0001 \msun ~ for stellar mass, 0.1 dex for helium layer masses, 0.1 for $X_o$, and 0.005 for $q_{\rm fm}$.

@@

In this study, we varied the parameters that have traditionally been varied in this type of asteroseismic fitting, so that we can place our results side by side with other studies of DBVs, including KIC 8626021 \citep{Bischoff-Kim14}, EC200058 \citep{Bischoff-Kim11a}, and CBS114 \citep{Metcalfe05a}. We now have 4 DBVs that were the object of asteroseismic fitting that used a consistent set of models. We find that cooler best fit models have thicker pure helium envelopes (Fig. \ref{ffit5}), in accordance with the outward diffusion of helium over time. To be consistent, we used the effective temperature found through the asteroseismic fitting of each star. Stellar mass also comes into play in diffusion. Ideally, we would like all 4 models to have the same mass. The models range in mass from 0.525 (EC20058) to 0.640 \msun ~ (CBS114). More detailed stellar evolution calculations would have to be performed in order to assess the significance of the trend observed with these 4 DBVs.

@@

In fitting GD358, we discovered that the best fit parameter space was semi-chaotic, with two very close sets of parameters able to yield very different fitness parameters. This is due to the fact that some periods can change values drastically for a small change in a given parameter. The phenomenon is only semi-chaotic, as the period will oscillate between two values. This is good news, because there remains a well defined minimum in parameter space, even if it is a double valued minimum. Any search algorithm that uses a small enough mesh should be able to find the minimum. This phenomenon likely became apparent while fitting GD358 because it is the first white dwarf for which we fit a long consecutive sequence of modes of the same $\ell$ with high resolution grids.

>>>>

\section{Astrophysical Context}
\label{intro}
White dwarfs are the end product of evolution for around 98\% of the stars in our Galaxy. 
Buried in their interiors are the records of physical processes that take place during 
earlier stages in the life of the star. Nuclear reaction rates during the core helium 
burning phase set the core composition of white dwarfs, while the relative time spent 
burning hydrogen and helium during the asymptotic-giant-branch (AGB) phase and mass-loss 
episodes determine the thickness of the helium layer \citep{Lawlor06,Althaus05}. Helium 
atmosphere white dwarfs (DBs) comprise roughly 20\% of the population of field white 
dwarfs, with most of the remaining 80\% consisting of their hydrogen atmosphere (DA) 
cousins. One theory behind the bifurcation into two spectral classes is that during post-AGB evolution, a very late thermal pulse can burn off the residual hydrogen 
in the envelope, producing a nearly pure helium atmosphere \citep{Iben83}. Such objects 
then return to the white dwarf cooling track as PG 1159 stars, which are 
widely believed to be the precursors of most DB white dwarfs. DBs are found to pulsate at 
effective temperatures ranging between 21,000 K and 28,000 K \citep{Beauchamp99, Castanheira05}.

The subject of this paper, GD358 (V777 Her) is the brightest ($m_{\rm v}=13.7$) and best 
studied helium atmosphere white dwarf pulsator. It is located near the red edge of the 
instability strip, with a spectroscopic temperature of \teff\ $=24000\pm500$ ~K and \logg\ $=7.8$ 
\citep{Koester2013}. GD358's pulsation spectrum contains a series of independent 
radial overtones, many with complex frequency structure.  For one epoch of data taken during a WET run, models involving magnetic fields and oblique rotation have been proposed to explain such structure \citep{Montgomery10}.

Since the 2006 Whole Earth Telescope (WET) run reported in \citet{Provencal09}, 
we have maintained an active observing program of GD358 in an effort to identify additional
frequencies in this complex star. These new observations have identified additional 
periods in GD358's pulsation spectrum, bringing the total known independent radial overtones to 15. Thirteen of 
these modes belong to a consecutive $\ell=1$ sequence, the longest sequence observed 
in a DBV.

Paradoxically, among the DBVs with enough detected periods to be fitted 
asteroseismically, GD358 is the only one that has not been analyzed using the complex 
C/O profiles adapted and parameterized from stellar evolution calculations
(e.g. {}\citet{Salaris97,Althaus05}). The most recent fits of GD358 \citep{Metcalfe03c} were 
performed using 11 observed modes and simple models where the oxygen abundance 
drops linearly from its central value to zero. We present here a new detailed asteroseismic analysis 
of GD358, taking advantage of its long sequence of $\ell=1$ modes to better constrain 
the asteroseismic fits and define the limits of high precision white dwarf asteroseismology.

@@

To compute our models, we used the White Dwarf Evolution Code (WDEC). The WDEC uses hot  and self-consistently allows them to relax to be solutions of the equations of stellar structure with the temperature of our choice.
polytrope models with temperatures above 100{,}000~K as starting model and numericaly evolves them until they are thermally relaxed 
solutions to the stellar structure equations and have the temperature of our choice. The mass is also fixed as an input, and so is the internal chemical composition profiles (no mass loss, no time dependent diffusion of elements). Each model we compute for our grids is the result 
of such an evolutionary sequence. The WDEC is described in detail in \citet{Lamb75} and 
\citet{Wood90}. We used smoother core composition profiles and implemented more complex profiles 
that result from stellar evolution calculations \citep{Salaris97}. We updated the envelope 
equation of state tables from those calculated by \citet{Fontaine77} to those given by 
\citet{Saumon95}. We use OPAL opacities \citep{Iglesias96} and plasmon neutrino rates 
published by \citet{Itoh96}.

@@

In our asteroseismic fits, we initially varied six parameters: the effective temperature, 
the mass and four structure parameters. There are two parameters associated with the shapes 
of the oxygen (and carbon) core composition profiles: the central oxygen abundance (\xo) and 
the edge of the homogeneous carbon and oxygen core (\qfm, as a fraction of stellar mass. 
For envelope structure, \menv\ marks the location of the base of the helium layer 
and \mhe\ the location where the helium abundance rises to $1$ (see Fig. \ref{ffit1}). 
\menv\ and \mhe\ are mass coordinates, defined as e.g. $M_{\rm env} = -\log(1 - M(r)/M_*)$, 
where $M(r)$ is the mass enclosed in radius r and $M_*$ is the stellar mass.

@@

\begin{equation}
\label{fiteq2}
\Delta P = \frac{\pi}{\sqrt{\ell(\ell + 1)}}\left[\int_{r_1}^{r_2}\frac{N}{r}dr\right]^{-1},
\end{equation}

@@

\begin{figure}
\epsscale{1.0}
\plotone{ffit3.eps}
%  \includegraphics[scale=0.70,viewport=50 0 50 350]{f2.png}
%  \includegraphics[width=0.9\columnwidth]{ffit3.eps}
\caption{
Dependence of the fitness parameter on different parameters. The 5 parameters other than the one shown 
on each plot are held to fixed values corresponding to the initial best fit model (table \ref{ffit1}). The vertical dashed lines mark the ranges considered in the refined fitting. For the envelope mass, we went from \menv = -2.0 to -2.3.
 \label{ffit3}
}
\end{figure}

@@

\subsection{Validation of the Fitting Method and Error Estimation}

@@

We first performed a fitting of the periods using the master (coarse) grid described in section \ref{grids}. This 
placed the best fit model in the appropriate region of parameter space. Then we refined our fits, using the grids 
described at the end of section \ref{periodspacing}. We are able to recover the best fit parameters adequately, with 
the all top 5 best fit models having parameters 
${\rm T_{eff}}=25000$~K, \menv $=-2.1$, \mhe $=-5.5$, $X_o=0.50$, and $q_{\rm fm}=0.190$ and a mass ranging between  0.5730 and 0.5734 \msun. The best fit model has \sigrms $=0.31$~s. We remind the reader that 
the step sizes in the second phase of the fitting are 50~K for ${\rm T_{eff}}$, 0.0001 \msun ~ for stellar mass, 0.1 dex for helium layer masses, 0.1 for $X_o$, and 0.005 for $q_{\rm fm}$.

This test also allows us to place minimum error bars on at least some of the parameters found: $\pm 600$ ~ K in effective temperature, 0.05 and 0.02 dex on \menv and \mhe respectively, 0.02 on $X_o$, and 0.002 on $q_{\rm fm}$.

@@

In this study, we varied the parameters that have traditionally been varied in this type of asteroseismic fitting, so that we can place our results side by side with other studies of DBVs, including KIC 8626021 \citep{Bischoff-Kim14,Giammichele17}, EC200058 \citep{Bischoff-Kim11a}, and CBS114 \citep{Metcalfe05a}. We now have 4 DBVs that were the object of asteroseismic fitting that used a consistent set of models. We find that cooler best fit models have thicker pure helium envelopes (Fig. \ref{ffit5}), in accordance with the outward diffusion of helium over time. To be consistent, we used the effective temperature found through the asteroseismic fitting of each star.

%There is one caveat to drawing conclusions from Fig. \ref{ffit5}. The initial amount of helium in the envelope depends on the mass of the progenitor star and its evolution, with more massive progenitors producing more massive white dwarfs according to the initial-final mass relation. Therefore, in order to isolate the effect of diffusion on the mass of the pure helium layer, one would have to consider a single white dwarf mass. However, the white dwarf mass distribution is known to be tight, with a range of progenitor masses yielding a narrow range of white dwarf masses. We do not believe that our precision in the asteroseismic analyses are at the level where we can use the relationship depicted in Fig. \ref{ffit5} to draw quantitative conclusions where the effect of white dwarf mass would matter.
%
%Stellar mass also comes into play in diffusion. Ideally, we would like all 4 models to have the same mass. The models range in mass from 0.525 (EC20058) to 0.640 \msun ~ (CBS114). More detailed stellar evolution calculations would have to be performed in order to assess the significance of the trend observed with these 4 DBVs.

@@

In fitting GD358, we discovered that two very close sets of parameters were able to yield very different fitness parameters. This is due to the fact that some periods can change values drastically for a small change in a given parameter. The periods tend to oscillate between two values and there remains a well defined minimum in parameter space, even if it is a double valued minimum. Any search algorithm that uses a small enough mesh should be able to find the minimum. This phenomenon likely became apparent while fitting GD358 because it is the first white dwarf for which we fit a long consecutive sequence of modes of the same $\ell$ with high resolution grids.
